Context-dependent consequences of lagged effects in demographic models

### 2024-09-03 docs/ipm_comparison_ms.qmd
back to it

---
title: "Context-dependent consequences of including lagged effects in demographic models"
shorttitle: "Lagged Effects IPMs"  
authors:
  - name: Eric R. Scott
    email: scottericr@gmail.com
    affil-id: 1
    corresponding: true
  - name: María Uriarte
    email: mu2126@columbia.edu
    institute: 2
  - name: Emilio M. Bruna
    email: embruna@ufl.edu
    affil-id: 1,3,5
affiliations:
  - id: "1"
    name: Department of Wildlife Ecology and Conservation, University of Florida, Gainesville, Florida 32611-0430 USA
  - id: "2" 
    name: Department of Ecology, Evolution and Environmental Biology, Columbia University 1200 Amsterdam Avenue, New York, New York 10027 USA
  - id: "3"
    name: Center for Latin American Studies, University of Florida, Gainesville, Florida 32611-5530 USA
  - id: "4"
    name: Biological Dynamics of Forest Fragments Project, INPA-PDBFF, CP 478, Manaus, Amazonas 69011-970 Brazil
date: 'last-modified'
date-format: '([draft:] D MMMM YYYY)'
linkcolor: blue
bibliography: references.bib
csl: "templates/the-american-naturalist.csl" # Insert path for the bib-style
# 
# highlights: |
#   These are the highlights
format: 
  pdf:
    number-sections: false
    colorlinks: true
    keeptex: true
    fontsize: 12pt
    documentclass: article
    # the following line is to render the aaffiliations in pdf, which is not default in quarto, see 
    # https://stackoverflow.com/questions/75040607/why-do-affiliations-not-show-up-anywhere-in-the-pdf-output-of-quarto
    template-partials:  
      - title.tex
    include-in-header:
      text: |
        \usepackage[noblocks]{authblk}
        \renewcommand*{\Authsep}{, }
        \renewcommand*{\Authand}{, }
        \renewcommand*{\Authands}{, }
        \renewcommand\Affilfont{\normalsize}
        \usepackage{lineno}\linenumbers
        \usepackage{sectsty}\sectionfont{\centering}
        \usepackage{sectsty}\subsectionfont{\centering}
        \usepackage{fontspec}
        \usepackage{titlesec}

@@

fig-cap-location: bottom
tbl-cap-location: top
geometry:
  - top=1in
  - left=1in
  - bottom=1in
  - right=1in
keywords: 
  - integral projection models
  - environmental stochasticity
  - lagged effects
---

<!-- This is the format for text comments that will be ignored during renderings. Do not put R code in these comments because it will not be ignored. -->

<!-- With the following code you can access and display values from the yml header above. -->

<!-- Keywords: `r rmarkdown::metadata$keywords` -->

<!-- Running title: `r rmarkdown::metadata$shorttitle` -->

<!-- Highlights: `r rmarkdown::metadata$highlights` -->

<!-- The following code chunk defines some general settings how code chunks should behave. -->

```{r setup, include=FALSE}
knitr::opts_chunk$set(
	echo = FALSE,
	fig.path = "figures/",
	dev = "png",
	dpi = 300,
	message = FALSE,
	warning = FALSE,
	collapse = TRUE,
	comment = "#>"
)
library(targets)
library(tidyverse)
library(kableExtra)
library(here)
```

@@

### **Paragraph 1**

1.  Structured demographic models, such as Integral Projection Models (IPMs) and matrix models, have emerged as invaluable tools in ecology, evolution, and conservation biology.

They provide insights into \_\_\_\_, the evolution of life-history strategies, and life-history stages and ecological processes at which conservation intervention is likely to have the greatest impact on population growth rate.

Their flexibility hasmade them increasingly applicable to species with differtent life histories.

however, there are still a number of challenges to address, namely \_\_\_\_ , \_\_\_\_, and \_\_\_\_.

### **Paragraph 2**

1.  It has long been recognized that there is the potential for lagged effects.
2.  Including lagged effects in models has been a major technical challenge, but there are now multiple approaches for doing so.
3.  The studies assessing the potential for lagged effects on vital rates find that they indeed appear to be prevalent [@eversLaggedDormantSeason2021; @scottDelayedEffectsClimate2022].

### **Paragraph 3**

1.  Lagged effects could be different in different habitat types.
2.  This could be a big reason for differences between habitats.

### **Paragraph 4**

1.  Here we...
2.  We conducted these analyses with both deterministic and stochastic IPMs.

@@

## Results

@@

@@

## Discussion

-   Our finding that the choice of IPM didn't change the relative ranking of CF and FF is consistent with @kayeEffectStochasticTechnique2003 finding that method effected stochastic lambda, but relative ranking of populations was consistent.

@@

@@

>>>>

---
title: "Context-dependent consequences of lagged effects in demographic models"
shorttitle: "Lagged Effects IPMs"  
authors:
  - name: Eric R. Scott
    email: scottericr@gmail.com
    affiliations:
      - ref: 1
    corresponding: true
  - name: María Uriarte
    email: mu2126@columbia.edu
    affiliations:
      - ref: 2
  - name: Emilio M. Bruna
    email: embruna@ufl.edu
    affiliations:
      - ref: 1
      - ref: 3
      - ref: 4
affiliations:
  - id: "1"
    name: Department of Wildlife Ecology and Conservation, University of Florida, Gainesville, Florida 32611-0430 USA
  - id: "2" 
    name: Department of Ecology, Evolution and Environmental Biology, Columbia University 1200 Amsterdam Avenue, New York, New York 10027 USA
  - id: "3"
    name: Center for Latin American Studies, University of Florida, Gainesville, Florida 32611-5530 USA
  - id: "4"
    name: Biological Dynamics of Forest Fragments Project, INPA-PDBFF, CP 478, Manaus, Amazonas 69011-970 Brazil
date: 'last-modified'
date-format: '([draft:] D MMMM YYYY)'
linkcolor: blue
bibliography: lagged-ipms-ms.bib
csl: "templates/the-american-naturalist.csl" # Insert path for the bib-style
# 
# highlights: |
#   These are the highlights
format: 
  pdf:
    number-sections: false
    colorlinks: true
    keeptex: true
    fontsize: 11pt
    documentclass: article
    # the following line is to render the aaffiliations in pdf, which is not default in quarto, see 
    # https://stackoverflow.com/questions/75040607/why-do-affiliations-not-show-up-anywhere-in-the-pdf-output-of-quarto
    template-partials:
      - title.tex
    include-in-header:
      text: |
        \usepackage[noblocks]{authblk}
        \renewcommand*{\Authsep}{, }
        \renewcommand*{\Authand}{, }
        \renewcommand*{\Authands}{, }
        \renewcommand\Affilfont{\normalsize}
        \usepackage{lineno}\linenumbers
        \usepackage{sectsty}\sectionfont{\centering}
        \usepackage{sectsty}\subsectionfont{\centering}
        \usepackage{fontspec}
        \usepackage{titlesec}
        \usepackage[sc]{mathpazo} %Like Palatino with extensive math support
        \usepackage[utf8]{inputenc}

@@

fig-cap-location: bottom
tbl-cap-location: top
geometry:
  - top=1in
  - left=1in
  - bottom=1in
  - right=1in
keywords: 
  - integral projection models
  - demography
  - population dynamics
  - environmental stochasticity
  - lagged effects

Notes: >
  Manuscript elements: Figure~1, figure~2, table~1, appendix~A.  
  Manuscript type: e-note

# abstract: >
#  The abstract should have 150 words. 
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
	echo = FALSE,
	fig.path = "figures/",
	dev = "png",
	dpi = 300,
	message = FALSE,
	warning = FALSE,
	collapse = TRUE,
	comment = "#>"
)
library(targets)
library(grateful)
library(tidyverse)
library(kableExtra)
library(here)
```

@@

### Paragraph 1: MM and IPM broadly used, but technical challenges remain

1.  Two types of structured population models - matrix models [@leslie1945use; @caswell2001] and integral projection models [@ellnerIntegralProjectionModels2006] - are fundamental frameworks used to study demography and population dynamics.\
2.  Their flexibility, in concert with a rapidly growing suite of software, data, and other resources [@ellnerDatadrivenModellingStructured2016; @levinIpmrFlexibleImplementation2021; @salguero-gomezCOMPADREPlantMatrix2015], have facilitated their use to study a wide range of topics in ecology, evolution, and conservation [@ellnerDatadrivenModellingStructured2016; @morrisQuantitativeConservationBiology2002; @croneHowPlantEcologists2011].
3.  While recent mathematical and statistical developments have expanded the scope of these applications [e.g., @brooksStatisticalModelingPatterns2019; @reesEvolvingIntegralProjection2016], a number of technical challenges remain.
4.  While many of these relate to general issues surrounding model development and validation [e.g., @williams2012avoiding], others stem from the fundamental challenge of using mathematical models - albeit highly advanced ones - to describe and analyze complex biological systems.

### Paragraph 2: Challenge of lagged effects

1.  Once such biological challenge is the implementation of alagged effects. It has long been recognized that there is the potential for lagged effects.
2.  Including lagged effects in models has been a major technical challenge, but there are now multiple approaches for doing so.
3.  The studies assessing the potential for lagged effects on vital rates find that they indeed appear to be prevalent [@eversLaggedDormantSeason2021; @scottDelayedEffectsClimate2022].

### Paragraph 3: Why does it matter when modeling?

1.  Lagged effects could be different in different habitat types.
2.  This could be a big reason for differences between habitats.
3.  Could also be important for stochastic vs. deterministic models.
4.  Here we...
5.  We conducted these analyses with both deterministic and stochastic IPMs.

@@

## Results & Discussion

@@

We found that the the choice of Integral Projection Model didn't change the relative ranking of lambda in Continous Forest and Fragments.

@@

These results are consistent with those of @kayeEffectStochasticTechnique2003, who found who found that the method effected stochastic lambda but relative ranking of populations was consistent.

@@

<!-- ```{r } -->

<!-- pkgs <- cite_packages(output = "table", out.dir = ".") -->

<!-- knitr::kable(pkgs) -->

<!-- ``` -->

@@

\`\`\`\`
